---
title: "calhoun-zach-homework5"
author: "Zach Calhoun"
date: "1/31/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## From the lab...
*This is for lab 1, first...*
```{r}
set.seed(1)

fx <- function(x) {
  return(sin(pi*x)^2)
}

x <- seq(0, 1, by=0.01)

plot(fx, xlim=c(0,1), ylim=c(0,1.5), ylab="f(x)", lwd=2)
curve(dunif, add = TRUE, col= "blue", lwd=2)
curve(dbeta(x,2,2), add=TRUE, col="red", lwd=2)
curve(0.7*dbeta(x,2,2), add=TRUE, col="green", lwd=2)
legend("bottom", legend=c(expression(paste("sin(",pi,"x)"^"2")), "Unif(0,1)","Beta(2,2)", "c*Beta(2,2)"), col=c("black","blue","red","green"), lty=c(1,1,1,1), bty= "n", cex=1.1, lwd=2)

```

## Task 4

```{r}
sim_fun <- function(f, envelope = "unif", par1 = 0, par2 = 1, n = 10^2, plot = TRUE){
  
  # Store functions for generating a random sample and
  # for calculating the likelihood of that sample
  r_envelope <- match.fun(paste0("r", envelope))
  d_envelope <- match.fun(paste0("d", envelope))
  proposal <- r_envelope(n, par1, par2)
  density_ratio <- f(proposal) / d_envelope(proposal, par1, par2)
  
  # Should this be changed? It seems like this should be changed
  # so the acceptance is only kept if the value is less than the
  # density ratio.
  samples <- proposal[runif(n) < density_ratio]
  acceptance_ratio <- length(samples) / n
  if (plot) {
    hist(samples, probability = TRUE, 
         main = paste0("Histogram of ", 
                       n, " samples from ", 
                       envelope, "(", par1, ",", par2,
                       ").\n Acceptance ratio: ",
                       round(acceptance_ratio,2)), 
                       cex.main = 0.75)
  }
  list(x = samples, acceptance_ratio = acceptance_ratio)
}

par(mfrow = c(2,2), mar = rep(4, 4))
unif_1 <- sim_fun(fx, envelope = "unif", par1 = 0, par2 = 1, n = 10^2) 
unif_2 <- sim_fun(fx, envelope = "unif", par1 = 0, par2 = 1, n = 10^5)
# ATTN: You will need to add in the Beta(2,2) densities on your own to finish task 4.
beta_1 <- sim_fun(fx, envelope = "beta", par1 = 2, par2 = 2, n = 10^2)
beta_1 <- sim_fun(fx, envelope = "beta", par1 = 2, par2 = 2, n = 10^5)

```


As shown in the graphs above, the accentance ratio ends up being the same for both the uniform and the beta distribution. This can be rationalized by the fact that the area between both of these curves and the pdf for which we are estimating appear to be roughly the same, which means that proposed samples will be rejected at the same rate using both proposals. In both cases, the acceptance ratio is less with fewer samples. This is caused by random noise, since the acceptance ratio converges to its "true" value when increasing the number of samples.

## Task 5

\textbf{i}: Without changing the constant $c$ applied to both of the proposal enveloping functions, the performance of both is about the same.

\textbf{ii}: However, Beta(2,2) ends up being the better enveloping function when we apply a constant to decrease the area between the curve and the complicated pdf. This can be demonstrated by looking at the graph below:

```{r}

set.seed(1)

fx <- function(x) {
  return(sin(pi*x)^2)
}

x <- seq(0, 1, by=0.01)

plot(fx, xlim=c(0,1), ylim=c(0,1.5), ylab="f(x)", lwd=2)
curve(dunif, add = TRUE, col= "blue", lwd=2)
curve(dbeta(x,2,2), add=TRUE, col="red", lwd=2)
curve(0.7*dbeta(x,2,2), add=TRUE, col="green", lwd=2)
legend("bottom", legend=c(expression(paste("sin(",pi,"x)"^"2")), "Unif(0,1)","Beta(2,2)", "c*Beta(2,2)"), col=c("black","blue","red","green"), lty=c(1,1,1,1), bty= "n", cex=1.1, lwd=2)

```

In this case, I applied $c=0.7$ to the Beta function to get the green curve, which is much closer to the complicated pdf we are trying to estimate, while still being greater than the sin function for all $X$. Because the area is decreased, we should expect that the acceptance ratio will be higher, as demonstrated below:

```{r}
# Update this function to include a constant C for improving the estimate.
sim_fun <- function(f, c=1, envelope = "unif", par1 = 0, par2 = 1, n = 10^2, plot = TRUE){
  
  # Store functions for generating a random sample and
  # for calculating the likelihood of that sample
  r_envelope <- match.fun(paste0("r", envelope))
  
  # Add in the C multiplier
  d_envelope <- match.fun(paste0("d", envelope))
  proposal <- r_envelope(n, par1, par2)
  density_ratio <- f(proposal) / d_envelope(proposal, par1, par2)
  
  # Should this be changed? It seems like this should be changed
  # so the acceptance is only kept if the value is less than the
  # density ratio.
  samples <- proposal[runif(n,0,c) < density_ratio]
  acceptance_ratio <- length(samples) / n
  if (plot) {
    hist(samples, probability = TRUE, 
         main = paste0("Histogram of ", 
                       n, " samples from ", 
                       envelope, "(", par1, ",", par2,
                       ").\n Acceptance ratio: ",
                       round(acceptance_ratio,2)), 
                       cex.main = 0.75)
  }
  list(x = samples, acceptance_ratio = acceptance_ratio)
}

par(mfrow = c(2,2), mar = rep(4, 4))
beta_1 <- sim_fun(fx, c=0.9, envelope = "beta", par1 = 2, par2 = 2, n = 10^2)
beta_1 <- sim_fun(fx, c=0.9, envelope = "beta", par1 = 2, par2 = 2, n = 10^5)
# ATTN: You will need to add in the Beta(2,2) densities on your own to finish task 4.
beta_1 <- sim_fun(fx, c=0.7, envelope = "beta", par1 = 2, par2 = 2, n = 10^2)
beta_1 <- sim_fun(fx, c=0.7, envelope = "beta", par1 = 2, par2 = 2, n = 10^5)


```

The two rows above show using a value of $c=0.9$ and $c=0.7$, respectively. We can see that the acceptance ratio is increased to 0.72 when using $c=0.7$, so we have a better performing sampler using that value of $c$.





